---
title: "Ormstrup Sø Dashboard"
logo: vt_cef.jpg
fig-width: 10
fig-asp: 0.3
format:
  dashboard:
      theme: 
        - brand
        - style.scss
      orientation: rows
      expandable: true
      scrolling: true
      nav-buttons:
      - icon: github
        href: https://github.com/FLARE-forecast/ORMS-forecast
      css: style.css
---

```{r}
#| label: set-inputs
library(tidyverse)
library(arrow)
library(bslib)
library(bsicons)
library(leaflet)
library(oce)

build_day <- Sys.Date() 

timezone <- 'Europe/Paris'

model_identifier <- 'flare_glm_v4_gefs_seamless'
interest_site <- 'ORMS'

### READ IN INSITU TARGETS
lake_directory <- getwd()
options(timeout=300)

Sys.setenv("AWS_REQUEST_CHECKSUM_CALCULATION" = "when_required",
           "AWS_RESPONSE_CHECKSUM_VALIDATION" = "when_required")

insitu_obs_url<- "https://amnh1.osn.mghpcc.org/bio230121-bucket01/flare/targets/ORMS/ORMS-targets-insitu.csv"

obs_cleaned <- readr::read_csv(insitu_obs_url) |> 
  filter(variable %in% c('temperature')) |> #, # was TEMP for CANN
         #depth == 0.5) |> ##insitu sensors and met 
  #dplyr::filter(is.na(inflow_name)) |> 
  collect() |> 
  lubridate::with_tz(datetime, tzone = timezone)

obs_updated <- max(obs_cleaned$datetime,na.rm = TRUE)


forecast_schema <- schema(
  reference_datetime = timestamp(unit = "us", timezone = "UTC"),
  datetime            = timestamp(unit = "us", timezone = "UTC"),
  pub_datetime         = timestamp(unit = "us", timezone = "UTC"),
  depth                = float64(),
  family               = utf8(),
  parameter            = utf8(),
  variable             = utf8(),
  prediction           = float64(),
  forecast             = float64(),
  variable_type        = utf8(),
  log_weight           = float64(),
  model_id             = utf8(),
  reference_date       = utf8()
)

s3 <- arrow::s3_bucket('bio230121-bucket01/flare/forecasts/parquet/site_id=ORMS', endpoint_override = 'amnh1.osn.mghpcc.org', anonymous = TRUE)

most_recent <-  arrow::open_dataset(s3, schema = forecast_schema) |> 
  filter(model_id == model_identifier) |>
  summarize(max = max(reference_datetime)) |> 
  collect() |> 
  pull()

current_time <- round_date(Sys.time(), unit = "minute")
current_time <- with_tz(current_time, tzone = "Europe/Paris")
```
::: {.callout-note collapse="true"}
## Disclaimer

The dashboard and all contents thereof and all services are provided on an "as is" basis without any warranty or condition, express, implied or statutory. User acknowledges any forecast is an experimental research product and there are uncertainties inherent in attempting to make projections, forecasts, and plans. User takes full responsibility for making its own evaluation of the adequacy and accuracy of all projections, forecasts, and plans
:::


Welcome to the the Ormstrup Sø (Lake Ormstrup) water temperature forecast dashboard. Forecasts are generated by the Virginia Tech Center for Ecosystem Forecasting ([CEF](https://www.ecoforecast.centers.vt.edu/)) in collaboration with Aarhus University Institute of Ecoscience partners.

THIS FORECAST IS ONLY A PROOF OF CONCEPT AND SHOULD NOT BE USED FOR DECISION MAKING OR ANY ACCESSMENT OF FORECAST QUALITY

Page last updated on `r current_time` local time.

## Most recent forecast

```{r}
## WATER TEMPERATURE
interest_var <- 'temperature'
interest_site <- 'ORMS'
day_range <- 14
focal_depths <- c(0.5, 5.0)

forecast <- arrow::open_dataset(s3, schema = forecast_schema) |> 
  filter(reference_datetime == most_recent,
         variable == "temperature",
         depth %in% c(focal_depths)) |> 
  collect() |> 
  lubridate::with_tz(datetime, tzone = "Europe/Paris") |> 
  lubridate::with_tz(reference_datetime, tzone = "Europe/Paris") |> 
  mutate(depth_labels = paste0("Depth: ", depth))

datetimes <- round_date(seq(min(forecast$datetime), max(forecast$datetime), by = "3 days"), unit = "day")
all_datetimes <- round_date(seq(min(forecast$datetime), max(forecast$datetime), by = "1 days"), unit = "day")
all_datetimes <- all_datetimes[!(all_datetimes %in% datetimes)]

ggplot(forecast, aes(x = datetime, y = prediction, group = parameter)) +
  geom_line() +
  geom_vline(aes(xintercept = reference_datetime)) +
  facet_wrap(~depth_labels) +
  theme_bw() +
  labs(y = "Water temperature (°C)") +
  scale_x_datetime(breaks = datetimes, minor_breaks = all_datetimes) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Probability of Mixing

```{r}



forecast <- arrow::open_dataset(s3, schema = forecast_schema) |> 
  mutate(parameter = as.character(parameter)) |> 
  filter(reference_datetime == most_recent,
         variable == "prob_mixed_density") |> 
  collect() |> 
  lubridate::with_tz(datetime, tzone = "Europe/Paris") |> 
  lubridate::with_tz(reference_datetime, tzone = "Europe/Paris")

ggplot(forecast, aes(x = datetime, y = prediction)) +
  geom_line() +
  geom_vline(aes(xintercept = reference_datetime)) +
  theme_bw() +
  labs(title = "Probability that lake is mixed", y = "Probability") +
  ylim(0,1.0) +
  scale_x_datetime(breaks = datetimes, minor_breaks = all_datetimes) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
## Location

```{r, include=FALSE}
sites <- suppressMessages(sf::st_read("sites.json"))
```

```{r fig.height=5, fig.width=5, echo=FALSE, include=TRUE, message = FALSE}
#sites <- suppressMessages(sf::st_read("sites.json"))
leaflet() %>%
  setView(lat = 56.326096, lng = 9.639093, zoom= 13) %>%
  addTiles(group="OSM") %>%
  addProviderTiles(providers$Esri.WorldImagery, group="Imagery") %>%
  addProviderTiles(providers$Esri.WorldTopoMap, group="Topo Map") %>%
  addLayersControl(baseGroups=c('Imagery','OSM', 'Topo Map')) |> 
  addMarkers(data  = sites, popup=~as.character(site_id), group = ~as.character(Partner))#, clusterOptions = markerClusterOptions())
```

## Learn more

More information about the project


::: {.card title = "Relevant Research from the Virginia Tech Center for Ecosystem Forecasting"}

Learn more about other forecasting applications by the Virginia Tech Center for Ecosystem Forecasting [here](https://www.ltreb-reservoirs.org/products/#publications-forecasting-and-modeling).

Learn more about the ongoing research at the Center for Ecosystem Forecasting [here](https://ecoforecast.centers.vt.edu/research.html).

:::
