---
title: "Ormstrup Sø Water Quality Forecast Dashboard"
logo: vt_cef.jpg
fig-width: 10
fig-asp: 0.3
format:
  dashboard:
      theme:
        - brand
        - style.scss
      orientation: rows
      expandable: true
      scrolling: true
      nav-buttons:
      - icon: github
        href: https://github.com/FLARE-forecast/ORMS-forecast
      css: style.css
---

```{r}
#| label: set-inputs
library(tidyverse)
library(arrow)
library(bslib)
library(bsicons)
library(leaflet)
library(oce)
library(plotly)
library(crosstalk)

build_day <- Sys.Date()

timezone <- 'Europe/Paris'

model_identifiers <- c('flare_glm_v4_gefs_seamless','flare_glm_v4_ecmwf_ifs025')
model_labels <- c(
  flare_glm_v4_gefs_seamless = "FLARE + NCEP GEFS",
  flare_glm_v4_ecmwf_ifs025  = "FLARE + ECMWF IFS"
)
model_colors <- c(
  "FLARE + NCEP GEFS" = "#51C2B9",
  "FLARE + ECMWF IFS" = "#CD4528"
)
relabel_model <- function(x) {
  labeled <- model_labels[x]
  ifelse(is.na(labeled), x, unname(labeled))
}

interest_site <- 'ORMS'
focal_depths <- c(0.5, 5.0)

### READ IN INSITU TARGETS
lake_directory <- getwd()
options(timeout=300)

Sys.setenv("AWS_REQUEST_CHECKSUM_CALCULATION" = "when_required",
           "AWS_RESPONSE_CHECKSUM_VALIDATION" = "when_required")

insitu_obs_url<- "https://amnh1.osn.mghpcc.org/bio230121-bucket01/flare/targets/ORMS/ORMS-targets-insitu.csv"

obs_cleaned <- readr::read_csv(insitu_obs_url) |>
  filter(variable %in% c('temperature')) |>
  collect() |>
  lubridate::with_tz(datetime, tzone = timezone)

obs_updated <- max(obs_cleaned$datetime,na.rm = TRUE)

forecast_schema <- arrow::schema(
  reference_datetime = timestamp(unit = "us", timezone = "UTC"),
  datetime            = timestamp(unit = "us", timezone = "UTC"),
  pub_datetime         = timestamp(unit = "us", timezone = "UTC"),
  depth                = float64(),
  family               = utf8(),
  parameter            = utf8(),
  variable             = utf8(),
  prediction           = float64(),
  forecast             = float64(),
  variable_type        = utf8(),
  log_weight           = float64(),
  model_id             = utf8(),
  reference_date       = utf8()
)

s3 <- arrow::s3_bucket('bio230121-bucket01/flare/forecasts/parquet/site_id=ORMS', endpoint_override = 'amnh1.osn.mghpcc.org', anonymous = TRUE)

most_recent <-  arrow::open_dataset(s3, schema = forecast_schema) |>
  filter(model_id %in% model_identifiers) |>
  summarize(max = max(reference_datetime)) |>
  collect() |>
  pull()

current_time <- round_date(Sys.time(), unit = "minute")
current_time <- with_tz(current_time, tzone = "Europe/Paris")

## PULL FORECAST DATA (shared by the KPI boxes and both charts below)
temp_forecast <- arrow::open_dataset(s3, schema = forecast_schema) |>
  filter(reference_datetime == most_recent,
         model_id %in% model_identifiers,
         variable == "temperature",
         depth %in% focal_depths) |>
  collect() |>
  lubridate::with_tz(datetime, tzone = "Europe/Paris") |>
  lubridate::with_tz(reference_datetime, tzone = "Europe/Paris") |>
  mutate(depth_labels = paste0("Depth: ", depth, " m"),
         parameter = paste(model_id, parameter, sep = "_"),
         model_id = relabel_model(model_id))

mixing_forecast <- arrow::open_dataset(s3, schema = forecast_schema) |>
  mutate(parameter = as.character(parameter)) |>
  filter(reference_datetime == most_recent,
         variable == "prob_mixed_density",
         model_id %in% model_identifiers) |>
  collect() |>
  lubridate::with_tz(datetime, tzone = "Europe/Paris") |>
  lubridate::with_tz(reference_datetime, tzone = "Europe/Paris") |>
  mutate(model_id = relabel_model(model_id),)

datetimes <- round_date(seq(min(temp_forecast$datetime), max(temp_forecast$datetime), by = "3 days"), unit = "day")
all_datetimes <- round_date(seq(min(temp_forecast$datetime), max(temp_forecast$datetime), by = "1 days"), unit = "day")
all_datetimes <- all_datetimes[!(all_datetimes %in% datetimes)]

## KPI summary values (model's first forecast timestep = its post-data-assimilation nowcast)
kpi_issued <- format(temp_forecast$reference_datetime[1], "%b %d, %Y %H:%M %Z")
kpi_horizon_days <- round(as.numeric(difftime(max(temp_forecast$datetime), temp_forecast$reference_datetime[1], units = "days")))
kpi_surface_temp <- temp_forecast |>
  filter(depth == 0.5, datetime == reference_datetime) |>
  summarize(mean = mean(prediction, na.rm = TRUE),
            min = min(prediction, na.rm = TRUE),
            max = max(prediction, na.rm = TRUE))
kpi_prob_mixed <- mixing_forecast |>
  filter(datetime == reference_datetime) |>
  summarize(value = mean(prediction, na.rm = TRUE)) |>
  pull(value)

## Adds a "now" text label above each facet panel of a ggplotly object.
## (The vertical line itself comes from a native geom_vline() in the ggplot
## call below - plotly::layout(p, shapes = ...) does not merge shapes added
## after ggplotly() in this plotly version, only annotations do, so the line
## and the label are added via two different mechanisms.)
add_now_label <- function(p, now_time, label_y = 0.97) {
  layout_names <- names(p$x$layout)
  x_axes <- grep("^xaxis", layout_names, value = TRUE)
  if (length(x_axes) == 0) x_axes <- "xaxis"
  y_axes <- grep("^yaxis", layout_names, value = TRUE)

  ## ggplotly() renders this datetime x-axis as type "linear" (raw epoch
  ## seconds), not "date" - a date string passed as x here can't be parsed
  ## against a linear axis and silently falls back to position 0 (the left
  ## edge), so "now" must be given as a plain number in the same units.
  now_num <- as.numeric(now_time)

  annotations <- lapply(sub("^xaxis", "", x_axes), function(sfx) {
    yref <- if (paste0("yaxis", sfx) %in% y_axes) paste0("y", sfx, " domain") else "y domain"
    ## xshift nudges the label a few pixels clear of the vline regardless of
    ## the axis's data units, since a data-coordinate offset would need to be
    ## rescaled per-axis (the two charts have very different x-axis spans).
    list(x = now_num, y = label_y, xref = paste0("x", sfx), yref = yref,
         text = "now", showarrow = FALSE, xanchor = "left", yanchor = "top", xshift = 5,
         font = list(size = 13, color = "#231F20", family = "Poppins"))
  })
  plotly::layout(p, annotations = annotations)
}

## For faceted plots, ggplotly() places the x-axis title as a plain paper
## annotation instead of using the axis's own "automargin"-aware title, so it
## doesn't account for the extra height of 45-degree rotated tick labels and
## ends up overlapping them. Push the title (and the plot's bottom margin)
## down far enough to clear the rotated labels, and optionally add extra top
## margin to give the chart title more breathing room above the facet strips.
## This mutates p$x$layout directly (not via plotly::layout()) because
## layout() queues annotations to be concatenated at build time rather than
## editing this one in place, which would leave the axis title duplicated.
## legend_gap_px adds further bottom margin (without moving the x-axis title
## itself) so a horizontal legend below the title has more room to sit lower,
## away from the title, instead of just moving both down together.
## extra_left_px widens the left margin: the rotated y-axis title is also a
## paper annotation here, positioned using a left margin that R estimated at
## render time from its own font metrics for Poppins - since R typically
## doesn't have Poppins installed and substitutes a narrower fallback font
## for that measurement, the browser (which does have the real Poppins font
## loaded) ends up needing a couple more pixels than were reserved, clipping
## the title against the edge of the plot's SVG canvas without this cushion.
push_down_facet_xtitle <- function(p, extra_px = 25, extra_top_px = 0, legend_gap_px = 0, extra_left_px = 0) {
  annotations <- p$x$layout$annotations
  idx <- which(vapply(annotations, function(a) identical(a$yref, "paper") && identical(a$y, 0), logical(1)))
  for (i in idx) annotations[[i]]$yshift <- annotations[[i]]$yshift - extra_px
  p$x$layout$annotations <- annotations
  p$x$layout$margin$b <- p$x$layout$margin$b + extra_px + legend_gap_px
  p$x$layout$margin$t <- p$x$layout$margin$t + extra_top_px
  p$x$layout$margin$l <- p$x$layout$margin$l + extra_left_px
  p
}

## Fixed pixel font sizes/margins (set above for desktop) don't scale down
## with the plot's container the way CSS would, so on a narrow phone-width
## card the same absolute-size legend/axis text and long "2026-07-03"-style
## tick labels overlap each other badly. This re-lays-out the plot in the
## browser once its actual container width is known (and again on resize):
## below ~500px it shrinks text, stacks the legend vertically, and swaps in
## short "M/D" tick labels instead of full dates; above that it restores the
## normal desktop sizing/legend defined in the R code below.
make_responsive <- function(p) {
  htmlwidgets::onRender(p, "
    function(el, x) {
      var monthAbbrev = ['Jan','Feb','Mar','Apr','May','Jun','Jul','Aug','Sep','Oct','Nov','Dec'];
      function shortDate(v) {
        var d = new Date(v * 1000);
        return monthAbbrev[d.getMonth()] + ' ' + d.getDate();
      }
      var lastMobile = null;
      function apply() {
        var mobile = el.clientWidth < 500;
        // Calling Plotly.relayout() unconditionally on every resize event
        // feeds back into bslib's own ResizeObserver-driven auto-sizing for
        // fill-container widgets (relayout changes the SVG's rendered size,
        // which retriggers bslib's resize handler, which can retrigger this
        // listener again), compounding the plot's height on each cycle.
        // Only relayout when the mobile/desktop breakpoint actually flips.
        if (mobile === lastMobile) return;
        lastMobile = mobile;
        var update = mobile ? {
          'legend.orientation': 'v',
          'legend.x': 0,
          'legend.xanchor': 'left',
          'legend.y': -0.15,
          'legend.font.size': 10,
          'legend.title.font.size': 10
        } : {
          'legend.orientation': 'h',
          'legend.y': -0.22,
          'legend.font.size': 13,
          'legend.title.font.size': 13
        };
        var tickSize = mobile ? 9 : 13;
        var titleSize = mobile ? 12 : 13;
        // A stacked (ncol=1) mobile facet layout gets its own x/y axis per
        // panel (xaxis2, yaxis2, ...), unlike the side-by-side desktop
        // layout where panels share one y-axis - looping over every axis
        // key found, rather than hardcoding 'xaxis'/'yaxis', is what makes
        // this apply consistently to every panel instead of just the first.
        Object.keys(el.layout).filter(function(k) { return /^xaxis/.test(k) || /^yaxis/.test(k); }).forEach(function(k) {
          update[k + '.tickfont.size'] = tickSize;
          update[k + '.title.font.size'] = titleSize;
          // Only x-axes carry the 'M/D'-style date ticktext to shorten -
          // y-axes are numeric (temperature/probability), and running them
          // through shortDate() would misread those values as timestamps.
          var ax = el.layout[k];
          if (/^xaxis/.test(k) && ax && ax.tickvals) {
            if (!ax._origticktext) ax._origticktext = ax.ticktext.slice();
            update[k + '.ticktext'] = mobile ? ax.tickvals.map(shortDate) : ax._origticktext;
          }
        });
        if (el.layout.yaxis && el.layout.yaxis.title && el.layout.yaxis.title.text) {
          var yTitle = el.layout.yaxis.title;
          if (!yTitle._origtext) yTitle._origtext = yTitle.text;
          var shortTitle = yTitle._origtext
            .replace('Water temperature', 'Water temp.')
            .replace('Probability', 'Prob.');
          update['yaxis.title.text'] = mobile ? shortTitle : yTitle._origtext;
        }
        function shortenTitleText(t) {
          return t.replace('Water temperature', 'Temp').replace('Probability', 'Prob.');
        }
        // Faceted plots render the x/y axis titles and facet strip labels as
        // plain paper annotations (see push_down_facet_xtitle above) rather
        // than native axis titles - those go through ggplot's pt-to-px
        // conversion (~1.33x) while the elements set directly above are
        // given raw pixel sizes, so restoring an annotation's own original
        // size would leave it visibly bigger than everything else. Target
        // the exact same pixel size as the rest of the mobile/desktop set.
        var annFontSize = mobile ? 10 : 13;
        if (el.layout.annotations) {
          el.layout.annotations.forEach(function(a) {
            if (a.annotationType === 'axis' && a.textangle === -90 && a._origtext === undefined) {
              a._origtext = a.text;
            }
          });
          update['annotations'] = el.layout.annotations.map(function(a) {
            var copy = Object.assign({}, a);
            if (copy.font) {
              copy.font = Object.assign({}, copy.font, {size: annFontSize});
            }
            if (copy._origtext !== undefined) {
              copy.text = mobile ? shortenTitleText(copy._origtext) : copy._origtext;
            }
            return copy;
          });
        }
        Plotly.relayout(el, update);
      }
      function syncSize() {
        // The CSS above pins this widget to a fixed height on desktop/tablet
        // (see the comment there for why), but Plotly.js only knows to
        // redraw its SVG to match a new container size when explicitly told
        // to - CSS alone changes the box, not the rendered plot inside it.
        Plotly.Plots.resize(el);
      }
      apply();
      syncSize();
      setTimeout(function() { apply(); syncSize(); }, 0);
      window.addEventListener('resize', function() { apply(); syncSize(); });
    }
  ")
}
```
::: {.callout-note collapse="true"}
## Disclaimer

The dashboard and all contents thereof and all services are provided on an "as is" basis without any warranty or condition, express, implied or statutory. User acknowledges any forecast is an experimental research product and there are uncertainties inherent in attempting to make projections, forecasts, and plans. User takes full responsibility for making its own evaluation of the adequacy and accuracy of all projections, forecasts, and plans
:::


Welcome to the the Ormstrup Sø (Lake Ormstrup) water temperature forecast dashboard. Forecasts are generated by the [Virginia Tech Center for Ecosystem Forecasting](https://www.ecoforecast.centers.vt.edu/) in collaboration with [Aarhus University Institute of Ecoscience](https://ecos.au.dk/en/) partners.

THIS FORECAST IS ONLY A PROOF OF CONCEPT AND SHOULD NOT BE USED FOR DECISION MAKING OR ANY ACCESSMENT OF FORECAST QUALITY

## At a Glance {height=15%}

### Forecast issued

```{r}
value_box(
  title = "Forecast issued",
  value = kpi_issued,
  showcase = bs_icon("calendar-check"),
  theme = "primary"
)
```

### Forecast horizon

```{r}
value_box(
  title = "Forecast horizon",
  value = paste(kpi_horizon_days, "days"),
  showcase = bs_icon("graph-up-arrow"),
  theme = "secondary"
)
```

### Surface temp now

```{r}
value_box(
  title = "Surface temp now (0.5 m)",
  value = paste0(round(kpi_surface_temp$mean, 1), " °C"),
  paste0("Range: ", round(kpi_surface_temp$min, 1), "–", round(kpi_surface_temp$max, 1), " °C"),
  showcase = bs_icon("thermometer-half"),
  theme = "primary"
)
```

### Probability mixed now

```{r}
value_box(
  title = "Probability lake is mixed",
  value = paste0(round(kpi_prob_mixed * 100), "%"),
  showcase = bs_icon("droplet-half"),
  theme = "secondary"
)
```

## Charts {height=50%}

### Most recent forecast {width=60%}

::: {.card title="Most recent forecast"}

```{r}
## Two depth panels side by side get crushed to ~150px wide each on a phone,
## so build the temperature chart as a function of facet column count: the
## default (side by side) for desktop/tablet, and ncol=1 (stacked, each panel
## full width) for a second widget shown only on narrow screens (see the
## .desktop-chart/.mobile-chart CSS toggle in style.css).
build_temp_widget <- function(ncol_facets = NULL) {
  # 1. Map your Crosstalk key
  shared_df <- highlight_key(temp_forecast, ~parameter)
  
  gg_temp <- ggplot(shared_df, aes(x = datetime, y = prediction, group = parameter, color = model_id)) +
    # Ensure lines map cleanly to the crosstalk parameter group
    geom_line(alpha = 0.2, linewidth = 0.4) +
    
    # Explicitly reinforce the group aesthetic to bind the point hover tooltips to the crosstalk key
    geom_point(aes(
      group = parameter, 
      text = paste0(
        "Model: ", model_id,
        "<br>Date: ", format(datetime, "%b %d, %Y %H:%M"),
        "<br>Depth: ", depth, " m",
        "<br>Predicted: ", round(prediction, 2), " °C"
      )), size = 0.3, alpha = 0.2, show.legend = FALSE) +
    
    geom_vline(aes(xintercept = reference_datetime)) +
    facet_wrap(~depth_labels, ncol = ncol_facets) +
    theme_light() +
    labs(x = NULL, y = "Water temperature (°C)", color = "Model") +
    scale_color_manual(values = model_colors) +
    scale_x_datetime(breaks = datetimes, minor_breaks = all_datetimes, date_labels = "%b %d") +
    theme(text = element_text(family = "Poppins", size = 13),
          axis.text.x = element_text(angle = 45, hjust = 1),
          strip.text = element_text(size = 13, color = "white", 
                                    margin = margin(t = 1, b = 1, unit = "pt")),
          strip.background = element_rect(fill = "gray30", color = NA),
          strip.clip = "on",
          legend.position = "bottom",
          panel.background = element_rect(fill = "transparent", color = NA),
          plot.background = element_rect(fill = "transparent", color = NA),
          legend.background = element_rect(fill = "transparent", color = NA))

  ggplotly(gg_temp, tooltip = "text") |>
    # 2. Reconfigure highlight to work safely on both lines and markers
    highlight(
      on = "plotly_click",        # Changed to click since selection brush is hidden
      off = "plotly_doubleclick",
      opacityDim = 0.2,       
      selected = attrs_selected(
        opacity = 1,              # Removed marker = list(color) to avoid crashing line geometries
        line = list( color = "black")
      )
    ) |>
    push_down_facet_xtitle(extra_px = 40, extra_top_px = 20, legend_gap_px = 5, extra_left_px = 10) |>
    add_now_label(now_time = temp_forecast$reference_datetime[1]) |>
    layout(
      legend = list(orientation = "h", y = -0.22),
      paper_bgcolor = "transparent", 
      plot_bgcolor = "transparent",
      font = list(family = "Poppins"),
      hoverlabel = list(font = list(family = "Poppins"))
    ) |>
    config(displayModeBar = FALSE) |>
    make_responsive()
}
```

::: {.desktop-chart}

```{r}
build_temp_widget()
```

:::

::: {.mobile-chart}

```{r}
build_temp_widget(ncol_facets = 1)
```

:::

:::

### Probability of Mixing {width=40%}

::: {.card title="Probability of Mixing"}

```{r}
gg_mix <- ggplot(mixing_forecast, aes(x = datetime, y = prediction, color = model_id)) +
  geom_line(linewidth = 0.6) +
  ## see note above gg_temp's geom_point() - text must live on a point layer,
  ## not on geom_line() itself, or ggplotly breaks the line at every point
  geom_point(aes(text = paste0(
      "Model: ", model_id,
      "<br>Date: ", format(datetime, "%b %d, %Y %H:%M"),
      "<br>Probability mixed: ", scales::percent(prediction, accuracy = 1)
    )), size = 1.2, show.legend = FALSE) +
  geom_vline(aes(xintercept = reference_datetime)) +
  theme_light() +
  labs(x = NULL, y = "Probability", color = "Model") +
  scale_color_manual(values = model_colors) +
  scale_y_continuous(limits = c(0, 1.0), oob = scales::squish, labels = scales::percent) +
  scale_x_datetime(breaks = datetimes, minor_breaks = all_datetimes, date_labels = "%b %d") +
  theme(text = element_text(family = "Poppins", size = 13),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom",
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.background = element_rect(fill = "transparent", color = NA),
        legend.background = element_rect(fill = "transparent", color = NA))

ggplotly(gg_mix, tooltip = "text") |>
  push_down_facet_xtitle(extra_px = 0, legend_gap_px = 5) |>
  add_now_label(now_time = mixing_forecast$reference_datetime[1], label_y = 0.90) |>
  layout(legend = list(orientation = "h", y = -0.22),
         paper_bgcolor = "transparent", plot_bgcolor = "transparent",
         font = list(family = "Poppins"),
         hoverlabel = list(font = list(family = "Poppins"))) |>
  config(displayModeBar = FALSE) |>
  make_responsive()
```

:::

## More info {height=35%}

### Location {width=40%}

```{r, include=FALSE}
sites <- suppressMessages(sf::st_read("sites.json"))
```

```{r, echo=FALSE, message=FALSE}
leaflet() %>%
  setView(lat = 56.326096, lng = 9.639093, zoom= 13) %>%
  addTiles(group="OSM") %>%
  addProviderTiles(providers$Esri.WorldImagery, group="Imagery") %>%
  addProviderTiles(providers$Esri.WorldTopoMap, group="Topo Map") %>%
  addLayersControl(baseGroups=c('Imagery','OSM', 'Topo Map')) |>
  addMarkers(data  = sites, popup=~as.character(site_id), group = ~as.character(Partner))#, clusterOptions = markerClusterOptions())
```

### Learn more {width=60%}

::: {.card title = "How forecasts are created"}

Forecasts are generated using the Forecasting Lakes and Reservoir Ecosystem (FLARE) system.  FLARE combines data assimilation (Ensemble Kalman Filter), the General Lake Model (GLM), and numerical weather forecasts to generate ensemble water quality forecasts.  It provides depth resolved outputs that include uncertainty from weather forecasts (via ensemble weather forecasts), GLM parameters, and GLM numerical representation of the ecosystem.  The forecasts are automatically generated every day at 13:00 UTC using GitHub Actions.

The forecasts presented here are driven by two different numerical weather forecast models.  The NCEF GEFS (National Center for Environmental Prediction Global Ensemble Forecasting System) is a 31-ensemble member forecast at the 0.25° resolution.  The ECMWF IFS (European Centre for Medium-Range Weather Forecasts Integrated Forecasting System) is a 51-ensemble member forecast at the the 0.25° resolution. Both forecast products are accessed through <https://open-meteo.com>.

The probability of mixing is calculated by comparing the density at the top and bottom of the lake for each ensemble member and determining the percentage of ensemble members with a density difference of less than 0.1 kg/m3

:::


::: {.card title = "Relevant research from the Virginia Tech Center for Ecosystem Forecasting"}

Learn more about other forecasting applications by the Center [here](https://www.ltreb-reservoirs.org/products/#publications-forecasting-and-modeling).

Learn more about the ongoing research at the Center [here](https://ecoforecast.centers.vt.edu/research.html).

:::

::: {#last-updated-footer}
Page last updated on `r current_time` Central European Time.
:::

